A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-15 paper/paper.tex
Add more discussion

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity.
% described
% in \citet{siles2012pure}
Section~\ref{sec:extension} explains how we
generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

@@

\section{Related Work}
\label{sec:relatedwork}
\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that establish metatheoretic results strong enough to derive
consistency for dependently typed systems. Some of these techniques
are still in use in more recent works. For example, the technique for
proving strong normalization by \citet{geuvers1994short} is adapted by
\citet{moon2021graded} to show the same property to a dependently
typed system extended with modalities.

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
uses a Kripke-style logical relation parameterized over an
type-directed equivalence relation satisfying certain properties. We
are able to avoid the complexivity of a Kripke-style logical relation
by defining our logical relation in terms of untyped parallel
reduction, which does not impose any scoping constraint on the
terms. \citet{decagda} defines their logical relation using the
induction-recursion scheme, which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq using a
special way of encoding induction recursion. Their work further
extends the decidability of type conversion result from~\citet{decagda} to the decidability
type checking of a bidrectional type system.

@@

>>>>

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity.
Section~\ref{sec:extension} explains how we
generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

@@

\section{Related Work and Discussion}
\label{sec:relatedwork}
\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that establish metatheoretic results strong enough to derive
consistency for dependently typed systems. Some of these techniques
are still in use in more recent works. For example, the technique for
proving strong normalization by \citet{geuvers1994short} is adapted by
\citet{moon2021graded} to show the same property to a dependently
typed system extended with modalities.

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
uses a Kripke-style logical relation parameterized over an
type-directed equivalence relation satisfying certain
properties. Their logical relation is defined using the
induction-recursion scheme, which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq using a
special way of encoding induction recursion. Their work further
extends the decidability of type conversion result from~\citet{decagda} to the decidability
type checking of a bidrectional type system.

@@

% To end this section, we

The final question we want to address is: why is our proof,
even with the extension from Section~\ref{sec:extension}, so much
shorter than the proofs from \citet{decagda, nbeincoq,
  martin-lof-a-la-coq}?
First, unlike developments that mechanize the correctness of a clever
type conversion algorithm, we only show the existence of
normal form for open and closed terms and state our properties in
terms of the untyped small-step reduction relation. This removes a lot
of scaffolding in the specification of our type system and our logical
relation.

This contrasts our development with
\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
difficult property to prove due to the complexity of the equality
judgment used to define the logical relation. In our system, the
irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
direct consequence of Lemma~\ref{lemma:interppreservation}, which is
straightforward after we prove the classical results about parallel
reduction in Section~\ref{sec:spec}. In other words, we keep the
the logical relation itself minimal at the cost of extra lemmas that can be proven
through syntactic means independently from the logical relation.
For example, as commented near the end of Section~\ref{sec:extension},
our normalization property (Corollary~\ref{corollary:exbetanf})
% can induce a decision procedure for type conversion by reducing two
% terms to $\beta\eta$-normal form and check for their syntactic
% equivalence. This decision procedure would be inefficient compared to
% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
% to weak head normal and checks for the equivalence of subterms
% recursively
does not immediately induce an efficient decision procedure, but it is
possible to recover an efficient algorithm by reasoning about untyped
lambda terms.

% the concision of
% our proof can be largely attributed to the fact that we bake into the
% logical relation only the minimum amount of information that is the
% most conducive to the proof of the metatheoretic result.

Another key to our proof is the very early establishment of the confluence
property of the paralell reduction relation
(Lemma~\ref{lemma:pardiamond}). This property is used in
Lemma~\ref{lemma:interppreservation} to show that the interpretation
of a type is preserved under evaluation. That is, if $[[InterpR i A
S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
In a system with typed directed reduction, the confluence result is
harder to establish since confluence requires subject reduction, which
circularly depends on $\Pi$-injectivity, a consequence of
confluence~\citep{siles2012pure}. There are different workarounds to
this problem. \citet{siles2012pure} proposes a syntactic approach by
defining a type system where the confluence result is
directly provable and later show that the system of interest is
equivalent to the system with the confluence property.
\citet{decagda} uses a relational model and the relational counterpart
of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
model is parameterized by a typed indexed equivalence relation and
is reused to prove derive different properties in their
development. Another perhaps simpler approach is to simply
extend the judgmental equality with rules about
$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
reduction to be proven independently from confluence and
$\Pi$-injectivity can later be shown to be admissible. We find the
approach from \citep{weirich:systemd} the most lightweight as it
allows us to derive confluence early on even in systems with typed
conversion.

In cases where the judgmental equality does not leverage type information heavily
(e.g. $\eta$-law for the unit type), it might be possible to show that
$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
property holds (we believe this is likely true for both
\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
terms of untyped parallel reduction to show the decidability of type
conversion. Otherwise, to fully model the system, we need to bake
typing information into our logical relation and we need to redefine
our logical relation in Kripke-style so we can relax the scoping
constraint in order to prove adequacy, though a relational model would
still be unnecessary as long as the $\eta$ laws are baked into the
typed directed relation.

Since confluence plays such as key role in our proof, we are uncertain
how to generalize our technique to systems where confluence does not
hold, and we will investigate those scenarios in our future work.

% We are able to avoid the complexivity of a Kripke-style logical relation
% by defining our logical relation in terms of untyped parallel
% reduction, which does not impose any scoping constraint on the
% terms.

% Why is it that the justification of $\eta$ law
% requires a Kripke-style relational model in all three developments
% but not ours? Are we ``cheating'' by defining \lang{}'s equational
% theory using an untyped conversion rule? Instead of answering the
% questions one by one, we clarify the design of our type system and our
% proof technique so the answers to those questions are self-evident.

% The key to the shortness of
